\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-01-20 appendix.qmd
Add section for total number of bins tuning

In the step of addressing low-density hexagons, which can arise from sparsely represented data in certain regions, we employ a systematic strategy. The goal is to ensure a more comprehensive coverage of the data by removing hexagons with low data density. To achieve this, we initiate the process by identifying, for each hex bin, the six nearest hex bins based on an equal 2D distance metric. Following this, we calculate the mean density (see @eq-equationp3), as outlined in the equation:

@@

The standard count is derived from the number of observations in the hex bins (see @eq-equationp2). Next, we examine the distribution of mean densities across all hex bins and designate the first quartile as the benchmark value for removing low-density hexagons. Finally, hex bins with mean densities below this benchmark value are removed from consideration. This meticulous procedure ensures the elimination of regions with inadequate data density, allowing the focus to shift to areas with more significant data representation. The result is the preservation of the overall structure of the data in the low-dimensional space, as illustrated in @fig-bintorm (c).

@@

>>>>

Addressing low-density hexagons is a systematic process to handle sparsely represented data in certain regions. For each hex bin, we identify the six nearest hex bins using an equal 2D distance metric. Then, we calculate the mean density, as outlined in the equations:

$$
\text{standard count} = \frac{\text{count}}{\text{max count}} 
$${#eq-equationp2}

$$
\text{mean density} = \frac{\text{standard count}}{6} 
$${#eq-equationp3}

The standard count is derived from the number of observations in the hex bins. By examining the distribution of mean densities and designating the first quartile as the benchmark value, hex bins with mean densities below this benchmark are removed. This process ensures the elimination of regions with insufficient data density, focusing on areas with more significant data representation and preserving the overall structure in the low-dimensional space.

<!--
In the step of addressing low-density hexagons, which can arise from sparsely represented data in certain regions, we employ a systematic strategy. The goal is to ensure a more comprehensive coverage of the data by removing hexagons with low data density. To achieve this, we initiate the process by identifying, for each hex bin, the six nearest hex bins based on an equal 2D distance metric. Following this, we calculate the mean density (see @eq-equationp3), as outlined in the equation:

@@

The standard count is derived from the number of observations in the hex bins (see @eq-equationp2). Next, we examine the distribution of mean densities across all hex bins and designate the first quartile as the benchmark value for removing low-density hexagons. Finally, hex bins with mean densities below this benchmark value are removed from consideration. This meticulous procedure ensures the elimination of regions with inadequate data density, allowing the focus to shift to areas with more significant data representation. The result is the preservation of the overall structure of the data in the low-dimensional space, as illustrated in @fig-bintorm (c).
-->

@@

<!--The removal of long edges is a critical step to create a smoother representation  by iteratively eliminating hexagons with excessive distances between centroids. This process eliminates outliers and noise while preserving essential local relationships within the data. To achieve this, distances between vertices are sorted, and unique distance values are extracted. By setting a threshold based on the largest difference between consecutive distance values, long edges are identified and removed. This refinement step contributes to enhancing the quality of the triangular mesh, ensuring a more accurate representation of the data structure.
-->

### 2024-01-20 paper.qmd
Add section for total number of bins tuning

library(Rtsne)
library(umap)
library(phateR)
library(patchwork)
library(langevitour)
library(colorspace)

library(grid)

@@

```{r}
#| echo: false
#| label: fig-binsize
#| fig-pos: H
#| fig-cap: "Hexbin plots from different number of bins for the **UMAP** projections of **S-curve** data: (a) b = 32 (4, 8), s = 1.643542, (b) b = 264 (12, 22), s = 1.643542, and (c) b = 840 (21, 40), s = 1.643542. The hexbins are colored based on the density of points, with darker colors indicating higher point density and yellow color representing lower point density within each bin. Does a value of number of bins exist to effectively represent the low-dimensional data?"

@@

In our model tuning process, we strategically adjust three key parameters to optimize the performance and accuracy of our approach. They are (i) the total number of bins ($b$), (ii) a benchmark value to remove low-density hexagons, and (iii) a benchmark value to remove long edges.

@@

Adjusting the parameter $b_1$ provides control over the total number of bins $b$, allowing us to fine-tune the hexagonal grid.

@@

num_bins_vec <- 1:14 ## Number of bins along the x-axis

@@

```{r}
#| echo: false
#| fig-cap: Goodness of fit statistics from different NLDR techniques applied to training S-curve dataset. What is the best NLDR technique to represent the original data in 2D?
#| label: fig-diagnosticpltScurve
#| out-width: 100%

@@

#### Remove low-density hexagons

Addressing low-density hexagons is a systematic process to handle sparsely represented data in certain regions. For each hex bin, we identify the six nearest hex bins using an equal 2D distance metric. Then, we calculate the mean density, as outlined in the equations:

@@

The standard count is derived from the number of observations in the hex bins. By examining the distribution of mean densities and designating the first quartile as the benchmark value, hex bins with mean densities below this benchmark are removed. This process ensures the elimination of regions with insufficient data density, focusing on areas with more significant data representation and preserving the overall structure in the low-dimensional space.

@@

#### Remove long edges

The removal of long edges is a critical step to create a smoother representation  by iteratively eliminating hexagons with excessive distances between centroids. This process eliminates outliers and noise while preserving essential local relationships within the data. To achieve this, distances between vertices are sorted, and unique distance values are extracted. By setting a threshold based on the largest difference between consecutive distance values, long edges are identified and removed. This refinement step contributes to enhancing the quality of the triangular mesh, ensuring a more accurate representation of the data structure.

@@

#### Goodness of fit statistics

@@

```{r}
#| warning: false
#| echo: false
#| label: fig-modelScurve
#| fig-cap: Is there a best model to represent the original data in 2D space or are they all providing equivalent information?, (a) Model in the 2D space with tSNE (<https://youtu.be/uy9SkAo6gAE>), (b) Model in the 2D space with UMAP (<https://youtu.be/0MJDhHrh_Ug>), (c) Model in the 2D space with PHATE (<https://youtu.be/HbVv0uy0QWk>), (d) Model in the 2D space with TriMAP (<https://youtu.be/2OGwipAjzc8>), and (e) Model in the 2D space with PaCMAP (<https://youtu.be/pkokI8d-cBk>).
#| out-width: 100%

@@

>>>>

library(Rtsne)
library(umap)
library(phateR)
library(patchwork)
library(langevitour)
library(colorspace)
library(kableExtra)

library(grid)
library(xtable)

xtable2kable <- function(x) {
  out <- capture.output(print(x, table.placement = NULL))[-(1:2)]
  out <- paste(out, collapse = "\n")
  structure(out, format = "latex", class = "knitr_kable")
}

@@

@@

The performance and robustness of our model depend on three key parameters: (i) the total number of bins ($b$), (ii) a benchmark value used to remove low-density hexagons, and (iii) a benchmark value used to remove long edges. However, there is no analytical formula to calculate an appropriate value for these parameters (see Appendix for details on how default values are calculated). The selection of these parameter values depends on the model performance computed by Mean Squared Error (MSE) (see @sec-goodfit).

@@

The number of hexagonal bins in the hexagonal grid has a considerable impact on the construction of the model. This is because it is the initial step in building the 2D model. The hexagonal grid with the chosen total number of bins must be able to capture the structure of the NLDR data. If the number of bins is too low, the model may not be able to capture the structure of the NLDR data effectively (see @fig-binsize (a)), while if there are too many bins, it may result in over-fitting the individual points of the NLDR data (see @fig-binsize (c)). Therefore, it is important to determine an appropriate number of bins to build an effective model.

Typically, the MSE plot will have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope will gradually decline or level off, indicating that a higher number of bins generates a smaller error. This MSE plot is useful in determining the effective number of bins required to construct the 2D model. @fig-diagnosticpltScurve (b) has created a graph that shows how the Mean Squared Error (MSE) changes depending on the number of bins assigned in the hexagonal grid for various NLDR techniques applied to the S-curve data set. As per the rule of thumb mentioned earlier, the effective number of bins for each NLDR method is shown in the @tbl-msebinsnldr. Furthermore, @fig-modelScurve shows the 2D models constructed for each NLDR technique with the selected total number of bins and how the models look in high-D with original data.

```{r msebinsnldr}
#| echo: false
#| message: false
#| warning: false

num_bins_for_nldr_methods_df <- tibble::tibble(NLDR_technique = "tSNE", tot_bins = 8)

num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
  tibble::add_row(NLDR_technique = "UMAP", tot_bins = 6)

num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
  tibble::add_row(NLDR_technique = "PHATE", tot_bins = 8)

num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
  tibble::add_row(NLDR_technique = "TriMAP", tot_bins = 6)

num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
  tibble::add_row(NLDR_technique = "PaCMAP", tot_bins = 10)

# knitr::kable(
#   num_bins_for_nldr_methods_df,
#   col.names = c('NLDR technique', 'total number of bins'), format = "latex", 
#              caption = "Total number of bins suggested by rule of thumb for each NLDR technique applied to S-curve dataset."
# ) |> 
#   kable_styling(latex_options = c("striped", "HOLD_position"), position = "center", full_width = FALSE) |>
#   row_spec(0, bold = TRUE)
num_bins_for_nldr_methods_df$tot_bins <- as.character(num_bins_for_nldr_methods_df$tot_bins)
colnames(num_bins_for_nldr_methods_df) <- c('NLDR technique', 'total number of bins')
num_bins_for_nldr_methods_df <- tibble::remove_rownames(num_bins_for_nldr_methods_df)

xtable(num_bins_for_nldr_methods_df, caption = "Total number of bins suggested by rule of thumb for each NLDR technique applied to S-curve dataset.", label = "tbl-msebinsnldr") |>
  xtable2kable() |> 
  kable_styling(latex_options = c("striped", "HOLD_position"), position = "center", full_width = FALSE) |>
  row_spec(0, bold = TRUE)

<!--Adjusting the parameter $b_1$ provides control over the total number of bins $b$, allowing us to fine-tune the hexagonal grid.-->

```{r}
#| echo: false
#| label: fig-binsize
#| fig-pos: H
#| fig-cap: "Hexbin plots from different number of bins for the **UMAP** projections of **S-curve** data: (a) b = 32 (4, 8), s = 1.643542, (b) b = 264 (12, 22), s = 1.643542, and (c) b = 840 (21, 40), s = 1.643542. The hexbins are colored based on the density of points, with darker colors indicating higher point density and yellow color representing lower point density within each bin. What is the number of bins that would be effective in representing low-dimensional data?"

@@

num_bins_vec <- 1:13 ## Number of bins along the x-axis

@@

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from different NLDR techniques applied to training S-curve dataset. What is the most effective NLDR technique to represent the original data in a 2D space? What is the effective number of bins in each NLDR technique to create a 2D model?"
#| label: fig-diagnosticpltScurve
#| out-width: 100%

@@

#### Benchmark value to remove low-density hexagons

<!--Addressing low-density hexagons is a systematic process to handle sparsely represented data in certain regions. For each hex bin, we identify the six nearest hex bins using an equal 2D distance metric. Then, we calculate the mean density, as outlined in the equations:

@@

The standard count is derived from the number of observations in the hex bins. By examining the distribution of mean densities and designating the first quartile as the benchmark value, hex bins with mean densities below this benchmark are removed. This process ensures the elimination of regions with insufficient data density, focusing on areas with more significant data representation and preserving the overall structure in the low-dimensional space.-->

@@

#### Benchmark value to remove long edges

<!--The removal of long edges is a critical step to create a smoother representation  by iteratively eliminating hexagons with excessive distances between centroids. This process eliminates outliers and noise while preserving essential local relationships within the data. To achieve this, distances between vertices are sorted, and unique distance values are extracted. By setting a threshold based on the largest difference between consecutive distance values, long edges are identified and removed. This refinement step contributes to enhancing the quality of the triangular mesh, ensuring a more accurate representation of the data structure.-->

@@

#### Goodness of fit statistics {#sec-goodfit}

@@

```{r}
#| warning: false
#| echo: false
#| label: fig-modelScurve
#| fig-cap: Is there a good model to represent the original data in 2D space or are they all providing equivalent information?, (a) Model in the 2D space with tSNE (<https://youtu.be/uy9SkAo6gAE>), (b) Model in the 2D space with UMAP (<https://youtu.be/0MJDhHrh_Ug>), (c) Model in the 2D space with PHATE (<https://youtu.be/HbVv0uy0QWk>), (d) Model in the 2D space with TriMAP (<https://youtu.be/2OGwipAjzc8>), and (e) Model in the 2D space with PaCMAP (<https://youtu.be/pkokI8d-cBk>).
#| out-width: 100%

@@

We introduce a new tool to help to determine which method, which parameter choice provide the most useful representation of high-D data.

### 2024-01-20 paper.tex
Add section for total number of bins tuning

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Visualising How Non-linear Dimension Reduction Warps Your Data},
  pdfauthor={Jayani P.G. Lakshika; Dianne Cook; Paul Harrison; Michael Lydeamore; Thiyanga S. Talagala},
  pdfkeywords={high-dimensional data, dimension
reduction, triangulation, hexagonal binning, low-dimensional
manifold, manifold learning, tour, data vizualization},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced]}{\end{tcolorbox}}\fi

@@

\caption{\label{fig-binsize}Hexbin plots from different number of bins
for the \textbf{UMAP} projections of \textbf{S-curve} data: (a) b = 32
(4, 8), s = 1.643542, (b) b = 264 (12, 22), s = 1.643542, and (c) b =
840 (21, 40), s = 1.643542. The hexbins are colored based on the density
of points, with darker colors indicating higher point density and yellow
color representing lower point density within each bin. Does a value of
number of bins exist to effectively represent the low-dimensional data?}

@@

\caption{\label{fig-wkhighD}How the 2D model lift into high dimensions?
(a) visualize the points and the hexagonal bin centroids related 2nd and
15th hexagons, (b) visualization of the edge connected the 2nd and 15th
hexagons (colored in red) in the triangular mesh. A video of tour
animation is available at
\url{https://www.youtube.com/watch?v=vBvM30Plt24}.}

@@

In our model tuning process, we strategically adjust three key
parameters to optimize the performance and accuracy of our approach.
They are (i) the total number of bins (\(b\)), (ii) a benchmark value to
remove low-density hexagons, and (iii) a benchmark value to remove long
edges.

@@

Adjusting the parameter \(b_1\) provides control over the total number
of bins \(b\), allowing us to fine-tune the hexagonal grid.

@@

Addressing low-density hexagons is a systematic process to handle
sparsely represented data in certain regions. For each hex bin, we
identify the six nearest hex bins using an equal 2D distance metric.
Then, we calculate the mean density, as outlined in the equations:

\begin{equation}\protect\hypertarget{eq-equationp2}{}{
\text{standard count} = \frac{\text{count}}{\text{max count}} 
}\label{eq-equationp2}\end{equation}

\begin{equation}\protect\hypertarget{eq-equationp3}{}{
\text{mean density} = \frac{\text{standard count}}{6} 
}\label{eq-equationp3}\end{equation}

The standard count is derived from the number of observations in the hex
bins. By examining the distribution of mean densities and designating
the first quartile as the benchmark value, hex bins with mean densities
below this benchmark are removed. This process ensures the elimination
of regions with insufficient data density, focusing on areas with more
significant data representation and preserving the overall structure in
the low-dimensional space.

@@

The removal of long edges is a critical step to create a smoother
representation by iteratively eliminating hexagons with excessive
distances between centroids. This process eliminates outliers and noise
while preserving essential local relationships within the data. To
achieve this, distances between vertices are sorted, and unique distance
values are extracted. By setting a threshold based on the largest
difference between consecutive distance values, long edges are
identified and removed. This refinement step contributes to enhancing
the quality of the triangular mesh, ensuring a more accurate
representation of the data structure.

\begin{itemize}
\tightlist
\item
  Compute high-D distances (using averaged high-D data)
\item
  Fit a linear model using response as distance\_high\_D and predictor
  as distance\_2D
\item
  Filter the model residuals only for short edges
\item
  compute error for each edge connection as (residual for that
  point)/(total residual for all edges * number of short edges)
\end{itemize}

@@

\hypertarget{goodness-of-fit-statistics}{%
\subsubsection{Goodness of fit
statistics}\label{goodness-of-fit-statistics}}

@@

\caption{\label{fig-diagnosticpltScurve}Goodness of fit statistics from
different NLDR techniques applied to training S-curve dataset. What is
the best NLDR technique to represent the original data in 2D?}

@@

\caption{\label{fig-modelScurve}Is there a best model to represent the
original data in 2D space or are they all providing equivalent
information?, (a) Model in the 2D space with tSNE
(\url{https://youtu.be/uy9SkAo6gAE}), (b) Model in the 2D space with
UMAP (\url{https://youtu.be/0MJDhHrh_Ug}), (c) Model in the 2D space
with PHATE (\url{https://youtu.be/HbVv0uy0QWk}), (d) Model in the 2D
space with TriMAP (\url{https://youtu.be/2OGwipAjzc8}), and (e) Model in
the 2D space with PaCMAP (\url{https://youtu.be/pkokI8d-cBk}).}

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Visualising How Non-linear Dimension Reduction Warps Your Data},
  pdfauthor={Jayani P.G. Lakshika; Dianne Cook; Paul Harrison; Michael Lydeamore; Thiyanga S. Talagala},
  pdfkeywords={high-dimensional data, dimension
reduction, triangulation, hexagonal binning, low-dimensional
manifold, manifold learning, tour, data vizualization},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, sharp corners, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi

@@

@@

\caption{\label{fig-wkhighD}How the 2D model lift into high dimensions?
(a) visualize the points and the hexagonal bin centroids related 2nd and
15th hexagons, (b) visualization of the edge connected the 2nd and 15th
hexagons (colored in red) in the triangular mesh. A video of tour
animation is available at \url{https://youtu.be/ffWFVK1UTIc}.}

@@

The performance and robustness of our model depend on three key
parameters: (i) the total number of bins (\(b\)), (ii) a benchmark value
used to remove low-density hexagons, and (iii) a benchmark value used to
remove long edges. However, there is no analytical formula to calculate
an appropriate value for these parameters (see Appendix for details on
how default values are calculated). The selection of these parameter
values depends on the model performance computed by Mean Squared Error
(MSE) (see Section~\ref{sec-goodfit}).

@@

The number of hexagonal bins in the hexagonal grid has a considerable
impact on the construction of the model. This is because it is the
initial step in building the 2D model. The hexagonal grid with the
chosen total number of bins must be able to capture the structure of the
NLDR data. If the number of bins is too low, the model may not be able
to capture the structure of the NLDR data effectively (see
Figure~\ref{fig-binsize} (a)), while if there are too many bins, it may
result in over-fitting the individual points of the NLDR data (see
Figure~\ref{fig-binsize} (c)). Therefore, it is important to determine
an appropriate number of bins to build an effective model.

Typically, the MSE plot will have a steep slope at the beginning,
indicating that a smaller number of bins causes a larger amount of
error. Then, the slope will gradually decline or level off, indicating
that a higher number of bins generates a smaller error. This MSE plot is
useful in determining the effective number of bins required to construct
the 2D model. Figure~\ref{fig-diagnosticpltScurve} (b) has created a
graph that shows how the Mean Squared Error (MSE) changes depending on
the number of bins assigned in the hexagonal grid for various NLDR
techniques applied to the S-curve data set. As per the rule of thumb
mentioned earlier, the effective number of bins for each NLDR method is
shown in the \textbf{?@tbl-msebinsnldr}. Furthermore,
Figure~\ref{fig-modelScurve} shows the 2D models constructed for each
NLDR technique with the selected total number of bins and how the models
look in high-D with original data.

\begin{table}[H]
\centering
\begin{tabular}{rll}
  \hline
\textbf{} & \textbf{NLDR technique} & \textbf{total number of bins }\\ 
  \hline
\cellcolor{gray!6}{1} & \cellcolor{gray!6}{tSNE} & \cellcolor{gray!6}{8 }\\ 
  2 & UMAP & 6 \\ 
\cellcolor{gray!6}{  3} & \cellcolor{gray!6}{PHATE} & \cellcolor{gray!6}{8 }\\ 
  4 & TriMAP & 6 \\ 
\cellcolor{gray!6}{  5} & \cellcolor{gray!6}{PaCMAP} & \cellcolor{gray!6}{10 }\\ 
   \hline
\end{tabular}
\caption{Total number of bins suggested by rule of thumb for each NLDR technique applied to S-curve dataset.} 
\label{tbl-msebinsnldr}
\end{table}

\caption{\label{fig-binsize}Hexbin plots from different number of bins
for the \textbf{UMAP} projections of \textbf{S-curve} data: (a) b = 32
(4, 8), s = 1.643542, (b) b = 264 (12, 22), s = 1.643542, and (c) b =
840 (21, 40), s = 1.643542. The hexbins are colored based on the density
of points, with darker colors indicating higher point density and yellow
color representing lower point density within each bin. What is the
number of bins that would be effective in representing low-dimensional
data?}

\caption{\label{fig-diagnosticpltScurve}Goodness of fit statistics from
different NLDR techniques applied to training S-curve dataset. What is
the most effective NLDR technique to represent the original data in a 2D
space? What is the effective number of bins in each NLDR technique to
create a 2D model?}

@@

@@

@@

\hypertarget{sec-goodfit}{%
\subsubsection{Goodness of fit statistics}\label{sec-goodfit}}

@@

@@

\caption{\label{fig-modelScurve}Is there a good model to represent the
original data in 2D space or are they all providing equivalent
information?, (a) Model in the 2D space with tSNE
(\url{https://youtu.be/uy9SkAo6gAE}), (b) Model in the 2D space with
UMAP (\url{https://youtu.be/0MJDhHrh_Ug}), (c) Model in the 2D space
with PHATE (\url{https://youtu.be/HbVv0uy0QWk}), (d) Model in the 2D
space with TriMAP (\url{https://youtu.be/2OGwipAjzc8}), and (e) Model in
the 2D space with PaCMAP (\url{https://youtu.be/pkokI8d-cBk}).}
